---
title: "WTI Options Portfolio"
format: 
  html:
    embed-resources: true
editor: visual
---

# **WTI Options Portfolio: Full Greek P&L Attribution**

```{r, include = FALSE}

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

library(tidyverse)
library(quantmod)
library(gt)
library(RTL)
library(data.table)
library(viridis)
```

```{r}

Fo <- 73
rf <- 0.04
Maturity = 3/12

port_table <- tibble::tibble(Position = c("P1", "P2", "P3"),
                             Type = c("Put", "Put", "Call"),
                             Strike = c(70, 55, 80),
                             Maturity,
                             Vol = c(0.35, 0.42, 0.3),
                             Contracts = c(+2000, -2000, -2000))

price_table <- tibble::tibble(price = seq(from = 50, to = 100, by = 1))



port_table %>% gt() %>% 
  tab_header(title = "WTI Options Portfolio")

```

### Q1

```{r}

spot_step <- 1

s_down <- Fo - spot_step
s_up <- Fo + spot_step

```

```{r}

greek_table <- expand.grid(Position = port_table$Position, 
                           Spot = seq(from = 48, to = 102, by = spot_step)) %>% 
  dplyr::left_join(port_table, by = join_by(Position)) %>% 
  dplyr::arrange(Position, Spot) %>% 
  rowwise() %>% 
  dplyr::mutate(
    Maturity = 3/12,
    Price = RTL::GBSOption(Spot,Strike,Maturity,rf,0, Vol, 
                           ifelse(Type == "Call", "call", "put"))$price,
    Pos_Price = Price*Contracts*1000) %>% 
  dplyr::group_by(Position) %>% 
  dplyr::mutate(Delta = (lead(Price)-lag(Price))/2*spot_step,
                Delta_Dol = Delta*Contracts*1000,
                Gamma = (lead(Price) - 2*Price + lag(Price))/spot_step^2,
                Gamma_Dol = Gamma*Contracts*1000) %>% 
  dplyr::filter(Spot %in% seq(from = 50, to = 100, by = 1))


```

```{r}
delta_p1 <- greek_table$Delta[which(greek_table$Position == "P1" &
                                            greek_table$Spot == Fo)]

delta_p2 <- greek_table$Delta[which(greek_table$Position == "P2" &
                                            greek_table$Spot == Fo)]

delta_p3 <- greek_table$Delta[which(greek_table$Position == "P3" &
                                            greek_table$Spot == Fo)]

gamma_p1 <- greek_table$Gamma[which(greek_table$Position == "P1" &
                                            greek_table$Spot == Fo)]

gamma_p2 <- greek_table$Gamma[which(greek_table$Position == "P2" &
                                            greek_table$Spot == Fo)]

gamma_p3 <- greek_table$Gamma[which(greek_table$Position == "P3" &
                                            greek_table$Spot == Fo)]
```

```{R}

delta_summary <- tibble::tibble(Position = c("P1", "P2", "P3"),
                               Delta = c(delta_p1, delta_p2, delta_p3))
                  
delta_summary %>% 
  gt() %>% 
    tab_header(title = "Position Deltas") %>% 
    cols_align(align = "center") %>% 
    tab_options(column_labels.background.color = "lightgrey",
                heading.background.color = "navy",
                column_labels.font.weight = "bold")

```

```{r}

greek_table %>%  
  ggplot2::ggplot(aes(x = Spot, y = Delta_Dol, color = Position))+
  geom_line(linewidth = .8)+
  facet_grid(Position~.,
             scales = "free_y")+
  theme_bw()+
  theme(
    element_rect(colour = "white"))+
  labs(y = "$ Delta")




```

```{r}
gamma_summary <- tibble::tibble(Position = c("P1", "P2", "P3"),
                               Gamma = c(gamma_p1, gamma_p2, gamma_p3))


gamma_summary %>% 
  gt() %>% 
    tab_header(title = "Position Gammas") %>% 
    cols_align(align = "center") %>% 
    tab_options(column_labels.background.color = "lightgrey",
                heading.background.color = "navy",
                column_labels.font.weight = "bold")

```

```{r}

greek_table %>%  
  ggplot2::ggplot(aes(x = Spot, y = Gamma_Dol, color = Position))+
  geom_line(linewidth = .8)+
  facet_grid(Position~.,
             scales = "free_y")+
  theme_bw()+
  theme(
    element_rect(colour = "white"))+
  labs(y = "$ Gamma")

```

```{r}

vol_step <- 0.01

vega_table <- expand.grid(Position = port_table$Position,
                          Spot = seq(from = 50, to = 100, by = 1),
                          i_vol = round(seq(from = 0.01, to = 1.00, by = vol_step),digits = 2)
                          ) %>% 
  dplyr::left_join(select(port_table, -Vol), by = join_by(Position)) %>% 
  dplyr::arrange(Position, i_vol, Spot) %>% 
  rowwise() %>% 
  dplyr::mutate(
    Price = RTL::GBSOption(Spot, 
                           Strike, 
                           Maturity, 
                           rf, 
                           0, 
                           i_vol, 
                           ifelse(Type == "Call", "call", "put"))$price) %>% 
  dplyr::group_by(Position) %>% 
  dplyr::mutate(Vega = ((lead(Price)-lag(Price))/2*spot_step)/100,
                 Vega_Dol = Vega*Contracts) %>%
  drop_na() %>% 
  ungroup()



```

```{r}

#p1_vol <- port_table$Vol[which(port_table$Position == "P1")]

p1_vol <- port_table %>% 
  filter(Position == "P1") %>% 
  pull(Vol)


p2_vol <- port_table$Vol[which(port_table$Position == "P2")]

p3_vol <- port_table$Vol[which(port_table$Position == "P3")]

vega_p1 <- vega_table$Vega[
  which(
    vega_table$Position == "P1" &
    vega_table$i_vol <= p1_vol+0.001 &
    vega_table$i_vol >= p1_vol-0.001 &
    vega_table$Spot == Fo)]


vega_p2 <- vega_table$Vega[
  which(
    vega_table$Position == "P2" &
    vega_table$i_vol <= p2_vol+0.001 &
    vega_table$i_vol >= p2_vol-0.001 &
    vega_table$Spot == Fo)]

vega_p3 <- vega_table$Vega[
  which(
    vega_table$Position == "P3" &
    vega_table$i_vol <= p3_vol+0.001 &
    vega_table$i_vol >= p3_vol-0.001 &
      vega_table$Spot == Fo)]







```

```{r}

vega_summary <- tibble::tibble(Position = c("P1", "P2", "P3"),
                               Vega = c(vega_p1, vega_p2, vega_p3)) 
                  

vega_summary %>% 
  gt() %>% 
    tab_header(title = "Position Vegas") %>% 
    cols_align(align = "center") %>% 
    tab_options(column_labels.background.color = "lightgrey",
                heading.background.color = "navy",
                column_labels.font.weight = "bold")

```

```{r}

vega_table %>%
  dplyr::filter(Spot %in% c(51:99)) %>% 
  dplyr::slice_head(n = -2) %>% 
  dplyr::filter(is.na(Vega) == FALSE) %>% 
  ggplot2::ggplot(aes(x = Spot, y = Vega))+
  geom_point(size = .8,
             aes(colour = i_vol))+
  facet_grid(Position~.,
             scales = "free_y")+
  labs(x = "Spot")+
  theme_bw()+
  theme(element_rect(colour = "white"))+
  viridis::scale_colour_viridis(option = "turbo")

```

```{r}

time_step <- 1/365

theta_table <-expand.grid(Position = port_table$Position,
                          Spot = seq(from = 49, to = 101, by = 1),
                          TTM = round(seq(from = 0, to = 2, by = 1/12),digits = 4)) %>% 
                dplyr::left_join(port_table) %>%
                dplyr::arrange(Position, TTM, Spot) %>%
                rowwise() %>% 
                dplyr::mutate(Price = RTL::GBSOption(Spot,
                                                     Strike,
                                                     TTM,
                                                     rf,
                                                     0,
                                                     Vol)$price,
                              Diff = abs(Maturity - TTM)) %>% 
                ungroup() %>%
                dplyr::mutate(Price = ifelse(is.nan(Price), 0, Price)) %>% 
                dplyr::group_by(Position, TTM) %>% 
                dplyr::mutate(Theta = (lead(Price) - lag(Price))/2*time_step,
                              Theta_Dol = Theta*Contracts) %>% 
  drop_na() %>% 
  ungroup()
              


```

```{r}

theta_p1 <- theta_table$Theta[
  which(theta_table$Position == "P1" & 
          theta_table$Diff ==min(theta_table$Diff, na.rm = TRUE) &
          theta_table$Spot == Fo
        )]

theta_p2 <- theta_table$Theta[
  which(theta_table$Position == "P2" & 
          theta_table$Diff ==min(theta_table$Diff, na.rm = TRUE) &
          theta_table$Spot == Fo
        )]

theta_p3 <- 
  theta_table$Theta[
  which(theta_table$Position == "P3" & 
          theta_table$Diff ==min(theta_table$Diff, na.rm = TRUE) &
          theta_table$Spot == Fo
        )]
```

```{r}

theta_summary <- tibble::tibble(Position = c("P1", "P2", "P3"),
                               Theta = c(theta_p1, theta_p2, theta_p3))


theta_summary %>% 
  gt() %>% 
  tab_header(title = "Position Thetas") %>% 
  cols_align(align = "center") %>% 
  tab_options(column_labels.background.color = "lightgrey",
              heading.background.color = "navy",
              column_labels.font.weight = "bold")



```

```{r}

theta_table %>% 
  dplyr::filter(is.na(Theta) == FALSE) %>% 
  ggplot2::ggplot(aes(x = Spot, y = Theta))+
  geom_point(aes(colour = TTM),size = .8)+
  facet_grid(Position~.,
             scales = "free_y")+
  theme_bw()+
  theme(element_rect(colour = "white"))+
  scale_colour_viridis(option = "turbo")


```

### Q1.d

```{r}

port_stats <- greek_table %>%
      ungroup() %>% 
      dplyr::select(Spot, Delta, Delta_Dol, Gamma, Gamma_Dol, Pos_Price) %>% 
      dplyr::group_by(Spot) %>% 
      dplyr::summarise(Port_Value = sum(Pos_Price), 
                       Port_Gamma = sum(Gamma),
                       Port_Delta = sum(Delta),
                       Port_Gamma_Dol = sum(Gamma_Dol),
                       Port_Delta_Dol = sum(Delta_Dol))

port_stats %>%
  dplyr::select(Spot, Port_Value, Port_Gamma, Port_Gamma_Dol) %>% 
  tidyr::pivot_longer(cols = starts_with("Port"),
                      names_to = "Stat",
                      values_to = "Value") %>% 
  dplyr::group_by(Stat) %>% 
  ggplot2::ggplot(aes(x = Spot, y = Value, colour = Stat))+
  geom_line(linewidth = .8)+
  facet_grid(Stat~.,
             scales = "free_y")+
  theme_bw()+
  theme(element_rect(colour = "white"),
        legend.position = "right")




```

The portfolio is short WTI - PnL falls as the WTI futures price increases.

Portfolio value is negatively convex, broadly explaining the negative gamma values through our price scope Fo = (50:100).

Gamma rises as options approach at-the-money (ATM) and flattens deeper in-the-money (ITM) and deeper out-the-money (OTM). The rate of change of delta peaks ATM which falls between both extremes and at the strike.

At spot \~ 65, gamma is elevated between P1 strike = 70 and P2 strike = 55. The same behavior is not observed at P2 strike \~ 55 because the gamma of P1 and P3 are heavily decayed, effectively neutralizing gamma of P2 and pushing it down to \[ G~P2~ - G~P1~ - G~P3~ \]. The same is true at strike = 80. The portfolio gamma is lowest as spot approaches 100 and the positions move significantly deep OTM for P1 and ITM for P2 and P3; in all scenarios, the rate of change of delta flattens, lowering gamma.

For delta hedgers, rising gamma means larger trades in the underlying to balance PnL changes in the option portfolio.

### Q2

```{r}

vega_sum_2 <- vega_table %>% 
  dplyr::filter((vega_table$Position == "P1" & vega_table$i_vol == p1_vol)|
                  (vega_table$Position == "P2" & vega_table$i_vol == p2_vol)|
                  (vega_table$Position == "P3" & vega_table$i_vol == p3_vol)) %>% 
  dplyr::select(Position, Spot, Vega, Vega_Dol) %>% 
  tidyr::pivot_wider(names_from = Position,
                     values_from = c(Vega, Vega_Dol))

theta_sum_2 <- theta_table %>%
  dplyr::select(-Maturity) %>% 
  dplyr::filter(theta_table$TTM == Maturity) %>% 
  dplyr::select(Position, Spot, Theta, Theta_Dol) %>%
  tidyr::pivot_wider(names_from = Position,
                     values_from = c(Theta, Theta_Dol))

portfolio_greeks <- greek_table %>% 
  dplyr::select(Position, Spot, Delta, Delta_Dol, Gamma, Gamma_Dol) %>% 
  tidyr::pivot_wider(names_from = Position,
                     values_from = c(Delta,
                                     Delta_Dol,
                                     Gamma,
                                     Gamma_Dol)) %>% 
  dplyr::left_join(theta_sum_2, by = join_by(Spot)) %>% 
  dplyr::left_join(vega_sum_2, by = join_by(Spot)) %>% 
  dplyr::mutate(Port_Delta = Delta_P1 + Delta_P2 + Delta_P3,
                Port_Gamma = Gamma_P1 + Gamma_P2 + Gamma_P3,
                Port_Vega = Vega_P1 + Vega_P2 + Vega_P3,
                Port_Theta = Theta_P1 + Theta_P2 + Theta_P3)


portfolio_greeks %>% 
  dplyr::select(Spot, starts_with("Port")) %>% 
  gt::gt() %>% 
  tab_header(title = "Portfolio Greeks") %>% 
  tab_options(container.height = 400,
              column_labels.background.color = "lightgrey",
              heading.background.color = "navy",
              column_labels.font.weight = "bold")
  




```

The portfolio delta flips at a WTI futures price of \~75-76.

### Q3

```{r}
#attributed
new_spot <- 90

spot_ch <- new_spot - Fo
iv_jump <- 0.2
time_jump <- 1/365

Attributed_PnL <- portfolio_greeks %>% 
  dplyr::mutate(Delta_Att = (Delta_Dol_P1 + Delta_Dol_P2 + Delta_Dol_P3)*spot_ch,
                Gamma_Att = (Gamma_Dol_P1 + Gamma_Dol_P2 + Gamma_Dol_P3)*spot_ch,
                Vega_Att = (Vega_Dol_P1 + Vega_Dol_P2 + Vega_Dol_P3)*iv_jump,
                Theta_Att = (Theta_Dol_P2 + Theta_Dol_P2 + Theta_Dol_P3)*time_jump,
                PnL = Delta_Att + Gamma_Att + Vega_Att + Theta_Att) %>% 
  dplyr::filter(Spot == Fo) %>% 
  dplyr::pull(PnL)
```

```{r}
#actual

Acc_Table <- port_table %>% 
  rowwise() %>% 
  dplyr::mutate(Spot = new_spot,
                Vol = Vol + iv_jump,
                Maturity = Maturity - time_jump,
                Price = RTL::GBSOption(Spot, 
                                       Strike, 
                                       Maturity, 
                                       rf, 
                                       0, 
                                       Vol,
                                       ifelse(Type == "Call", "call", "put"))$price,
                Pos_Val = Price*Contracts*1000
                )

Actual_PnL <- sum(Acc_Table$Pos_Val)

Actual_PnL



```
